---
title: "Common raster operations using terra"
author: "Matt Williamson"
date: "9/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Your assignment

The goals for this assignment are to:

- practice raster operations using the `terra` package

- begin using functional programming for repetitive operations

By the time you work through these steps you should have a `SpatRaster` object with elevation, climate, and land value data that is cropped to Idaho.


## Step 1: Load your packages
We will rely on the `terra` and `geodata` packages for this assignment. Load them here
```{r packageload}

```


## Step 2: Use `geodata` to download some data

Our spatial analyses are often built upon a mix of data that we (or our colleagues) generated ourselves and data that is freely available through a variety of platforms and portals (see the Resources section of the course page if you're interested in a few other data options). In some cases, very nice people have generated `R` packages to facilitate data access _within_ the `R` script to aid in reproducibility. We'll be using the `geodata` package today, but there are a number of others (the [`FedData` ](https://github.com/ropensci/FedData) package is another handy one). Use the `elevation_30s()` and the `cmip6_world()` functions to download a raster of eleveation for the United States and the projected maximum temperature across the globe (you'll need to use the `?` to access the helpfiles to understand how these functions work). Make sure to assign the result of each to an object (i.e., one object in your workspace should contain the elevation data and one should contain the climate data). Inspect these objects.

```{r dldata}

```

* __Question 1:__ What class is each object returned by the functions? What are their origins, extents, and resolutions? What CRS are they using?
* __Question 2:__ Look at the helpfile for `rast()`. How might you determine the number of layers in the elevation and climate objects you created? 
* __Question 3:__ How many layers are in each object?
* __Question 4:__ How does the way that `terra` handles multiple raster layers differ from that of `raster`? (Hint: look up `?raster::`RasterBrick-class``)


## Step 3: Load your own data
Now that you know a bit more about the `rast()` function you should be able to use it to load some raster data from a file (rather than download it within a package). You'll find the land value raster from session 4 along with Dave Theobald's [human modification index](https://databasin.org/datasets/110a8b7e238444e2ad95b7c17e889b66/) raster in our shared folder (`/opt/data/session08/`). Load them here.

```{r loaddata}

```


## Step 4: Line up our rasters.
As we have discussed, before we can do much in terms of spatial analysis, we need to get all of our pieces to line up. This involves checking to make sure each raster has the same CRS, extent, origin, and resolution (note that having the same resolution is mostly a convenience for combining layers in a single object, changing resolution has a lot of consequences for our analyses so do it with some thought). We'll use `project()` and `crop()` to get the data cut down to size and then a combination of `aggregate()`, `disaggregate()`, and `resample()` to get the resolutions and origins aligned. Let's use the land value raster as our template raster (i.e., we eventually want all of our data to be cropped to the extent of land value dataset with the same origin, extent, and resolution). Remmeber, that it's usually easier to project the raster with the smaller extent to the appropriate CRS and then move forward with cropping, etc.

```{r alignrast}

```

* __Question 5:__ How did you decide whether to use `aggregate()`, `disaggregate()`, or `resample()`? Did any of the layers require a specific function?

## Step 4b: Try it again with an empty template raster
One of the nice things about how raster operations work in `R` is that you can move your data into new CRS's, extents, and resolutions even if you don't have a dataset that's currently in that form. This happens by using a blank template raster - one that has the desired extent, origin, CRS, and resolution, but maybe doesn't have values (this is how it's often done in the examples for the packate). Try that here - create a template raster with a 5km resolution, with an extent that is slightly smaller than the land value raster (see `?extent` for details on setting an extent object) and in an equal area projection of your choice. Then, Line up all of your rasters with this this new template raster. Do that here.

```{r aligntemp}

```


## Step 5: Let's run some focal statistics
Now that we've got our data aligned, Let's use a moving window to 'smooth' the data. Use the `focal()` function to take the mean of a 5x5 neighborhood and then a 10x10 neighborhood. You can do this for whichever raster you choose. _Bonus_ create a function that you apply to a multi-layer `SpatRast` object (see the `lapp` function in `terra`) 

```{r focalstats}

```

* __Question 6:__ Use raster math to generate the difference between the two smoothed outpots and plot the result.

## Bonus: Create a function that does all of this. 
Now that you've seen how tedious it can be to copy and paste things over and over. Try to create a single function that will take any list of `SpatRast` objects; crop, align, and resample them to a template raster; and smooth them. If you're really ambitioius use the examples [here](http://adv-r.had.co.nz/Profiling.html) to figure out which part of your code is the slowest.
